---
title: "Homework 8"
author: "Sucry Bendeck"
date: "last-modified"
format: html
editor: visual
embed-resources: true
code-tools: true
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(dplyr)
library(jsonlite)
library(tidyr)
library(vegabrite)
```

# Data

```{r}
ds<- fromJSON("https://calvin-data304.netlify.app/data/swd-lets-practice-ex-2-3.json")
head(ds)
```

# Exercise #1

## Graphic 1

```{r}

vl_chart() |>
  vl_mark_bar() |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_y("value", type = "quantitative") |>
  vl_encode_xOffset("type") |>
  vl_encode_color("type") |>
  vl_add_data(ds) |>
  vl_fold(c("capacity", "demand"), as = c("type", "value"))

```

## Graphic #2

```{r}
chart <- vl_chart(height = 400, width = 600) |>
  vl_mark_bar() |>
  vl_encode_x("date", type = "nominal", scale = list(padding = 0.3)) |>
  vl_add_data(ds)

vl_layer(
  chart |>
    vl_mark_bar() |>
    vl_encode_y("capacity"),
  
  chart |>
    vl_mark_bar() |>
    vl_encode_y("demand") |>
    vl_encode_size("demand", scale = list(range = c(5, 20)))
)

```

## Graphic #3
```{r}
base_chart <- vl_chart(height = 400, width = 600) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_color("type") |>
  vl_encode_detail("type") |>
  vl_add_data(ds)

vl_layer(
  base_chart |>
    vl_mark_point() |>
    vl_encode_y("capacity", type = "quantitative"),
  
  base_chart |>
    vl_mark_point() |>
    vl_encode_y("demand", type = "quantitative"),
  
  base_chart |>
    vl_mark_line() |>
    vl_encode_x("date", type = "nominal") |>
    vl_encode_y("capacity", type = "quantitative") |>
    vl_encode_y2("demand", type = "quantitative") |>
    vl_encode_color("type")
)
```

## Graphic #4

```{r}
 base_chart <- vl_chart(height = 400, width = 600) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_color("type:n") |>
  vl_add_data(ds)

vl_layer(
  base_chart |>
    vl_mark_line() |>
    vl_encode_y("capacity", type = "Q"),
  
  base_chart |>
    vl_mark_line() |>
    vl_encode_y("demand", type = "Q")
)


```

## Graphic #5
```{r}
capDem <- vl_chart(height = 400, width = 600) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_color("type") |>
  vl_add_data(ds)|>
  vl_fold(c("capacity", "demand"), as = c("type", "value"))

vl_layer(
  capDem |>
    vl_mark_bar() |>
    vl_encode_y("capacity", type = "quantitative") |>
    vl_encode_color("type"),
  
  capDem |>
    vl_mark_bar() |>
    vl_encode_y("unmet_demand", type = "quantitative") |>
    vl_encode_color("type") 
)


```


## Graphic #6
```{r}
# data wrangling

ds <- ds |>
  mutate(unmet_demand = demand - capacity)

# Graphic 5
vl_chart(height = 400, width = 600) |>
  vl_mark_line(point = TRUE) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_y("unmet_demand", type = "quantitative") |>
  vl_add_data(ds)

```

# Exercise #2

## Reading data:

```{r}
jobs<- fromJSON("https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json")
head(jobs)
```


## Graphic #1

```{r}
vl_chart()|>
vl_mark_point(tooltip=list(content = "data"))|>
  vl_pivot("year",groupby = list("job","sex"), value = "perc")|>
  vl_encode_x("1950:Q")|>
  vl_encode_y("2000:Q")|>
  vl_scale_x(type= "symlog",constant = 1)|>
    vl_scale_y(type= "symlog",constant = 1)|>
  vl_facet_column("sex",title = "")|>
  vl_encode_tooltip_array(list("job","sex","1850","2000"))|>
  vl_add_data(jobs)

```

## Graphic #2

```{r}
vl_chart()|>
  vl_mark_point(tooltip=list(content = "data"))|>
  vl_pivot("year",groupby = list("job","sex"), value = "perc")|>
  vl_encode_x("1950:Q")|>
  vl_encode_y("2000:Q")|>
  vl_scale_x(type= "symlog",constant = .00001)|>
    vl_scale_y(type= "symlog",constant = 0.00001)|>
  vl_facet_column("sex",title = "")|>
  vl_add_properties(title = list(text= "Percent of people working various jobs", 
                                 subtitle= "each dot represents one ocupation"))|>
   vl_encode_tooltip_array(list("job","sex","1850","1950"))|>
vl_add_data(jobs)

```
## *A:*

```{r} 
#Grouping the data
jobs_summary <- jobs |>
  group_by(year, sex) |>
  summarize(total_perc = sum(perc, na.rm = TRUE))
```

```{r}
#Stacked bar( easier to observe the total perc)
vl_chart(height = 400, width = 600) |>
  vl_mark_bar() |>
  vl_encode_x("year", type = "ordinal") |>
  vl_encode_y("total_perc", type = "quantitative", stack = "zero") |>
  vl_encode_color("sex") |>
  vl_add_data(jobs_summary)

# Line( easier to observe trend ovetr time)
vl_chart(height = 400, width = 600) |>
  vl_mark_line() |>
  vl_encode_x("year", type = "ordinal") |>
  vl_encode_y("total_perc", type = "quantitative", stack = "zero") |>
  vl_encode_xOffset("sex") |>
  vl_encode_color("sex") |>
  vl_add_data(jobs_summary)
```

The first graph appears to show the proportion of men and women in the total job market. This provides insight into the distribution of employment between men and women. The second graph focuses on the trend of employment share over time for both genders, revealing that the values for men and women are gradually converging toward 0.5. This suggests that gender participation in the workforce is becoming more equal over time.
## *B:*
```{r}
# Selecting Years
jobYears <- jobs |>
  filter(year == 1970 | year == 1980 | year == 1990 | year == 2000)

# Creating the chart
vl_chart(title = "") |>
  vl_mark_point() |>
  vl_encode_y("job", type = "nominal") |>
  vl_encode_x("perc", type = "quantitative") |>
  vl_encode_color("sex") |>
  vl_encode_shape("sex") |>
  vl_facet_column("year", title = "") |>
  vl_add_data(jobYears) |>
  vl_encode_text("perc", format = ".0%") # Format perc as percentage

```

# Exercise #3

```{r}
DataSet1<- read.csv("https://calvin-data304.netlify.app/data/pump_price_for_gasoline_us_per_liter.csv")

DataSet2<- read.csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv")
```

