---
title: "Homework 9"
author: "Sucry Bendeck"
date: "last-modified"
format: html
editor: visual
embed-resources: true
code-tools: true
---

# Exercise 1(Wilke)

## A. (Chapter 4):

I learned ,in class and in the reading, that color is a tool that can aid us to do multiple things in our graphics-things such as:

-   Represent different data points.
-   Highlight elements in the graphic.
-   Make the groups of data different.

## B. (Chapter 5):

The organization the Wilke gives the graphics is based on the what story is the author through the data. Wilke gives us six categories:

-   Amounts

-   Distribution

-   Proportions

-   x-y prop

-   Geospatial data\

-   Uncertainty I appreciate that Wilke tells us the difrent limitations that each of the graphics have. Wilke mentioned that certain graphics reduce the ambiguity of the data which and so by using certain marks you may help the audience have a better comprehension of the visualization.

## C. (Graphs that I dont know)

Here are a few of the graphics I dont know how to do:

-   Mosaic Plot

-   Tree Map

-   Parallel Sets

-   Density Contours

-   Hex Bins

-   Correlogram

-   Cartogram

-   Cartogram heatmap

-   Confidence Strips

-   Eyes

-   Half eyes

## D. (Graphics attempts)
```{r, include=FALSE}
library(dplyr)
library(jsonlite)
library(vegabrite)
library(tidyr)
library(sf)
```


```{r}

# Load data from the URL
data_url <- "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/cars.json"
data <- fromJSON(data_url)

# Create the chart
chart <- vl_chart(height = 400,width = 600) |>
  vl_mark_rect() |>
  vl_encode_x("Origin:N") |>
  vl_encode_y("Cylinders:N") |>
  vl_encode_size("count:Q") |>
  vl_encode_color("Origin:N",legend = FALSE) |>
  vl_config(
    mark = list(
      opacity = 0.7,
      strokeWidth = 0.5
    ),
    axis = list(
      domain = FALSE,
      ticks = FALSE,
      grid = FALSE
    )
  )|> 
  vl_add_data(data)

# Show the chart
chart
```

## E.(Chapters 6 - 16)

This tree map clearly easily shows the proportions of the car data which is what I am trying to do here. It also uses space effectivly by filling the area with the data.

# Exercise 2(More pie)

## A.

```{r}

ficData<-tribble(
  ~number, ~ response, ~count, ~year,
  5, "strongly agree", 230, "this year",
  4, "agree", 320, "this year",
  3, "neither agree nor disagree", 192, "this year",
  2, "disagree", 95, "this year",
  1, "strongly disagre", 157, "this year",
  0, "no reponse", 52, "this year",
  5, "strongly agree", 240, "last year",
  4, "agree", 275, "last year",
  3, "neither agree nor disagree", 182, "last year",
  2, "disagree", 125, "last year",
  1, "strongly disagre", 189, "last year",
  0, "no reponse", 32, "last year"
)

data_prop <- ficData|>
  group_by(year) |>
  mutate(proportion = count / sum(count))
# check data types
str(ficData)

pie_chart0 <- vl_chart(height=200, width = 300,data = data_prop) |>
  vl_encode_theta(field = "proportion", type = "quantitative") |>
  vl_encode_color(field = "response", type = "nominal") |>
  vl_encode_tooltip(field = "proportion", type = "quantitative") |>
  vl_facet(field = "year", type = "ordinal", title = "") |>
  vl_mark_arc()

# Render the pie chart
pie_chart0


```

```{r}
# graph
stacked_bar_chart <- vl_chart(width= 400, height = 200,data = data_prop) |>
  vl_mark_bar() |> 
  vl_encode_x(field = "year", type = "nominal") |>
  vl_encode_y(field = "proportion", type = "quantitative") |>
  vl_encode_color(field = "response", type = "nominal", scale = list(
    domain = c("strongly agree", "agree", "neither agree nor disagree", "disagree", "strongly disagree", "no response"),
    range = c("green", "lightgreen", "yellow", "orange", "red", "gray")
  ))

# Render the stacked bar chart
stacked_bar_chart

```

### pros and cons

It’s going to depend on the story you’re trying to tell. Pie charts are better for simple comparisons of proportions within a single category (like a year), however they can be less precise.Stacked bar charts are more effective if you want to observe trends across multiple years.

# Exercise 3 (Create a choropleth map)

## Map
### Data

```{r}
library(vegabrite)
library(tidyverse)
library(sf)

# Reading and cleaning the data
USAdams <- read.csv("usadams.csv") |> drop_na(longitude, latitude, id, name)
sampled_USAdams <- USAdams |> sample_n(1000)
# longitude and latitude to numeric
sampled_USAdams <- sampled_USAdams |> mutate(
  longitude = as.numeric(longitude),
  latitude = as.numeric(latitude)
)

#  no NAs
sampled_USAdams_clean <- sampled_USAdams |>
  filter(!is.na(longitude) & !is.na(latitude))

dam_points_sf <- st_as_sf(sampled_USAdams_clean, coords = c("longitude", "latitude"), crs = 4326)

# Transform the dam data to Albers USA projection (EPSG: 5070)
dam_points_albers <- st_transform(dam_points_sf, crs = 5070)

us_geojson_url <- "https://cdn.jsdelivr.net/npm/vega-datasets@v1.29.0/data/us-10m.json"
```



### US maps
```{r}

# U.S. map visualization
us_map <- vl_chart(width = 600, height = 400) |>
  vl_add_data(url = us_geojson_url,
              format = list(type = "topojson", feature = "states")) |>
  vl_mark_geoshape() |>
  vl_encode_color(value = "lightgray") |>
  vl_add_properties(projection = list(type = "albersUsa"))
us_map

```


```{r}
# Transform the dam data 
dam_points_albers <- st_transform(dam_points_sf, crs = 5070)

# Extract transformed coordinates
sampled_USAdams_clean_albers <- st_coordinates(dam_points_albers)

# Combine the transformed coordinates with the original data
sampled_USAdams_clean_albers <- cbind(sampled_USAdams_clean, sampled_USAdams_clean_albers)

# Get the min and max values of the transformed coordinates
x_range <- range(sampled_USAdams_clean_albers$X)
y_range <- range(sampled_USAdams_clean_albers$Y)

us_projection_x_range <- c(-2500000, 2500000) 
us_projection_y_range <- c(0, 4000000)  

scaled_x <- scales::rescale(sampled_USAdams_clean_albers$X, to = us_projection_x_range)
scaled_y <- scales::rescale(sampled_USAdams_clean_albers$Y, to = us_projection_y_range)

# Create the scatter plot with adjusted coordinates and the correct projection
scatter_plot <- vl_chart(data = sampled_USAdams_clean_albers, width = 600, height = 400) |>
  vl_mark_point() |>
  vl_encode_x("X", type = "quantitative", scale = list(domain = c(x_range[1], x_range[2]))) |>
  vl_encode_y("Y", type = "quantitative", scale = list(domain = c(y_range[1], y_range[2]))) |>
  vl_encode_size(value = 10) |>
  vl_encode_color(value = "red") |>
  vl_add_properties(
    projection = list(type = "albersUsa")  # Ensure same projection as the US map
  )

scatter_plot


```

```{r}

# Combine the U.S. map and the scatter plot with layering
overlay_map <- vl_layer(
  us_map,
  scatter_plot
)

overlay_map

```

