---
title: "Homework 8"
author: "Sucry Bendeck"
date: "last-modified"
format: html
editor: visual
embed-resources: true
code-tools: true
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(dplyr)
library(jsonlite)
library(tidyr)
library(vegabrite)
```

# Data

```{r}
#| code-fold: true
ds<- fromJSON("https://calvin-data304.netlify.app/data/swd-lets-practice-ex-2-3.json")
head(ds)
```

# Exercise #1

## Graphic 1

```{r}
#| code-fold: true

vl_chart() |>
  vl_mark_bar() |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_y("value", type = "quantitative") |>
  vl_encode_xOffset("type") |>
  vl_encode_color("type") |>
  vl_add_data(ds) |>
  vl_fold(c("capacity", "demand"), as = c("type", "value"))

```

## Graphic #2

```{r}
#| code-fold: true
chart <- vl_chart(height = 400, width = 600) |>
  vl_mark_bar() |>
  vl_encode_x("date", type = "nominal", scale = list(padding = 0.3)) |>
  vl_add_data(ds)

vl_layer(
  chart |>
    vl_mark_bar() |>
    vl_encode_y("capacity"),
  
  chart |>
    vl_mark_bar() |>
    vl_encode_y("demand") |>
    vl_encode_size("demand", scale = list(range = c(5, 20)))
)

```

## Graphic #3
```{r}
#| code-fold: true
base_chart <- vl_chart(height = 400, width = 600) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_color("type") |>
  vl_encode_detail("type") |>
  vl_add_data(ds)

vl_layer(
  base_chart |>
    vl_mark_point() |>
    vl_encode_y("capacity", type = "quantitative"),
  
  base_chart |>
    vl_mark_point() |>
    vl_encode_y("demand", type = "quantitative"),
  
  base_chart |>
    vl_mark_line() |>
    vl_encode_x("date", type = "nominal") |>
    vl_encode_y("capacity", type = "quantitative") |>
    vl_encode_y2("demand", type = "quantitative") |>
    vl_encode_color("type")
)
```

## Graphic #4

```{r}
#| code-fold: true
 base_chart <- vl_chart(height = 400, width = 600) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_color("type:n") |>
  vl_add_data(ds)

vl_layer(
  base_chart |>
    vl_mark_line() |>
    vl_encode_y("capacity", type = "Q"),
  
  base_chart |>
    vl_mark_line() |>
    vl_encode_y("demand", type = "Q")
)


```

## Graphic #5
```{r}
#| code-fold: true
capDem <- vl_chart(height = 400, width = 600) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_color("type") |>
  vl_add_data(ds)|>
  vl_fold(c("capacity", "demand"), as = c("type", "value"))

vl_layer(
  capDem |>
    vl_mark_bar() |>
    vl_encode_y("capacity", type = "quantitative") |>
    vl_encode_color("type"),
  
  capDem |>
    vl_mark_bar() |>
    vl_encode_y("unmet_demand", type = "quantitative") |>
    vl_encode_color("type") 
)


```


## Graphic #6
```{r}
#| code-fold: true
# data wrangling
ds <- ds |>
  mutate(unmet_demand = demand - capacity)

# Graphic 5
vl_chart(height = 400, width = 600) |>
  vl_mark_line(point = TRUE) |>
  vl_encode_x("date", type = "nominal") |>
  vl_encode_y("unmet_demand", type = "quantitative") |>
  vl_add_data(ds)

```

# Exercise #2

### Reading data:

```{r, include=FALSE}
#| code-fold: true
jobs<- fromJSON("https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json")
head(jobs)
```


## Graphic #1

```{r}
#| code-fold: true
vl_chart()|>
vl_mark_point(tooltip=list(content = "data"))|>
  vl_pivot("year",groupby = list("job","sex"), value = "perc")|>
  vl_encode_x("1950:Q")|>
  vl_encode_y("2000:Q")|>
  vl_scale_x(type= "symlog",constant = 1)|>
    vl_scale_y(type= "symlog",constant = 1)|>
  vl_facet_column("sex",title = "")|>
  vl_encode_tooltip_array(list("job","sex","1850","2000"))|>
  vl_add_data(jobs)

```

## Graphic #2

```{r}
#| code-fold: true
vl_chart()|>
  vl_mark_point(tooltip=list(content = "data"))|>
  vl_pivot("year",groupby = list("job","sex"), value = "perc")|>
  vl_encode_x("1950:Q")|>
  vl_encode_y("2000:Q")|>
  vl_scale_x(type= "symlog",constant = .00001)|>
    vl_scale_y(type= "symlog",constant = 0.00001)|>
  vl_facet_column("sex",title = "")|>
  vl_add_properties(title = list(text= "Percent of people working various jobs", 
                                 subtitle= "each dot represents one ocupation"))|>
   vl_encode_tooltip_array(list("job","sex","1850","1950"))|>
vl_add_data(jobs)

```
## *A:*

```{r} 
#| code-fold: true
#Grouping the data
jobs_summary <- jobs |>
  group_by(year, sex) |>
  summarize(total_perc = sum(perc, na.rm = TRUE))
```

```{r}
#| code-fold: true
#Stacked bar( easier to observe the total perc)
vl_chart(height = 400, width = 600) |>
  vl_mark_bar() |>
  vl_encode_x("year", type = "ordinal") |>
  vl_encode_y("total_perc", type = "quantitative", stack = "zero") |>
  vl_encode_color("sex") |>
  vl_add_data(jobs_summary)

# Line( easier to observe trend ovetr time)
vl_chart(height = 400, width = 600) |>
  vl_mark_line() |>
  vl_encode_x("year", type = "ordinal") |>
  vl_encode_y("total_perc", type = "quantitative", stack = "zero") |>
  vl_encode_xOffset("sex") |>
  vl_encode_color("sex") |>
  vl_add_data(jobs_summary)
```

The first graph appears to show the proportion of men and women in the total job market. This provides insight into the distribution of employment between men and women. The second graph focuses on the trend of employment share over time for both genders, revealing that the values for men and women are gradually converging toward 0.5. This suggests that gender participation in the workforce is becoming more equal over time.\

## *B:*
```{r}
#| code-fold: true
# Selecting Years
jobYears <- jobs |>
  filter(year == 1970 | year == 1980 | year == 1990 | year == 2000)

# Creating the chart
vl_chart(title = "") |>
  vl_mark_point() |>
  vl_encode_y("job", type = "nominal") |>
  vl_encode_x("perc", type = "quantitative") |>
  vl_encode_color("sex") |>
  vl_encode_shape("sex") |>
  vl_facet_column("year", title = "") |>
  vl_add_data(jobYears) |>
  vl_encode_text("perc", format = ".0%") # Format perc as percentage

```


```{r}
#| code-fold: true

vl_chart(title = "Employment Percentage by Job and Sex") |>
  vl_mark_bar() |>
  vl_encode_y("job", type = "nominal") |>
  vl_encode_x("perc", type = "quantitative", stack = "zero") |>
  vl_encode_color("sex") |>
  vl_facet_column("year") |>
  vl_encode_xOffset("sex")|>
  vl_add_data(jobYears) |>
  vl_facet_row("sex", title = "")|>
  vl_encode_text("perc", format = ".0%") # Format perc as percentage
```

# Exercise #3

```{r,include=FALSE}
#| code-fold: true
DataSet1<- read.csv("https://calvin-data304.netlify.app/data/pump_price_for_gasoline_us_per_liter.csv")

DataSet2<- read.csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv")
```

## Wrangling the data:
```{r, warning=FALSE}
#changing name to country so the merge is easier.
DataSet2<-DataSet2|> rename(country= name)

# Merging the datasets
 gasprice<-left_join(DataSet1,DataSet2, by = "country")

gasprice <- gasprice |>
  mutate(across(starts_with("X"), as.numeric))

# Pivot the data while keeping NAs
gasprice_long <- gasprice %>%
  pivot_longer(cols = starts_with("X"), names_to = "year", values_to = "price", values_drop_na = FALSE)
```

```{r, results='hide', echo = "FALSE", include=FALSE}

# Check for countries
unique(gasprice$country)

# Check for countries after pivot
unique(gasprice_long$country)
```

### Check countries  differences  
```{r}
#| code-fold: true
# Check countries differences 
setdiff(unique(gasprice$country), unique(gasprice_long$country))

```


```{r}
#| code-fold: true
# Pivot and filter the data for USA
gasprice_long_usa <- gasprice |>
  pivot_longer(cols = starts_with("X"), names_to = "year", values_to = "price") |>
 mutate(
    year = gsub("X", "", year),
    year = as.integer(year),
    price_per_gallon = price * 3.79 # Convert price from per liter to per gallon
  ) |>
  filter(country == "USA")

# Create the chart
vl_chart(title = "Gas Prices Trend in the USA by Year") |>
  vl_mark_line(point = TRUE) |>
  vl_encode_x("year", type = "nominal",title = "Year") |>
  vl_encode_y("price_per_gallon", type = "quantitative",title = "US $ per gallon") |>
  vl_add_data(gasprice_long_usa)

```



